Reintroduction of resistant frogs facilitates landscape-scale recovery in the presence of a lethal fungal disease

### 2023-12-21 doc/manuscript/translocation_bioarxiv_SI.tex
Minor edits to Intro and Results

Swab extracts were analyzed using standard Bd DNA extraction and qPCR
methods \citep{boyle2004}, and extracts were analyzed singly instead of
in triplicate \citep{kriger2006}. For analysis of swabs collected during
2005--2014, we used standards developed from known concentrations of
zoospores \citep{boyle2004}, and after 2014, we used standards based on
single ITS1 PCR amplicons \citep{longo2013}. Based on paired comparisons
between samples analyzed using both types of standards, Bd in the study
area has an average of 60 ITS1 copies per zoospore. To express all qPCR
results as the number of ITS1 copies, starting quantities obtained using
the zoospore standard (measured as ``zoospore equivalents'') were
multiplied by 60. In addition, all qPCR quantities (regardless of
standard) were multiplied by 80 to account for the fact that DNA
extracts from swabs were diluted 80-fold during extraction and PCR
\citep{vredenburg2010}.

@@

We estimated survival and recruitment for each site using open
population CMR models based on \citep{joseph2018}. For each individual
\(i=1, ..., M\) on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if
the individual was not detected, and \(o_{i, j}=2\) if the individual
was detected. Capture histories of \(M\) individuals are modeled, though
only \(N_s\) individuals were captured.

This parameter expanded data augmentation allows us to capture the
possibility that undetected individuals may have recruited into the
adult population \citep{royle2012}. Here, \(M\) was chosen to be three
times the number of observed individuals (\(3N_s\)) to be considerably
greater than our prior guess of \(N_s\).

@@

We computed the likelihood of each individual capture history using the
forward algorithm, and we estimated the latent states using the
forward-backward algorithm \citep{zucchini2009, joseph2018}.

All of the code to specify and fit the model in Stan is available in the
open source mrmr package \citep{joseph2019}.

@@

We used Bayesian generalized linear mixed models to investigate
predictors of survival among sites. The response \(y_i\) is binary,
representing a point estimate of whether individual \(i\) survived in
the year following translocation. We generated these point estimates by
rounding the posterior median of 1-year post-introduction survival for
each individual, and modeled the data using a Bernoulli distribution:

@@

These models were fit using the \texttt{stan\_glmer()} function in the
rstanarm package, with default priors described below
\citep{rstanarm2022}. These priors are vague, but include data dependent
scaling as follows to account for different input variable scales.
However, as we standardized all predictor variables similarly to have
equal variance (by centering and dividing by twice the sample standard
deviation), the resulting priors are identical. Specifically, we have:

@@

The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
rstanarm is a zero-mean Gaussian, where the covariance matrix is
constructed from a correlation matrix with an LKJ prior, and a vector of
variance parameters -- the decomposition of variance prior with unit
regularization, concentration, shape, and scale parameters
\citep{lewandowski2009}.

We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
with four parallel chains, each run for 10,000 iterations, discarding
the first half of each chain as warm-up draws \citep{rstanarm2022}. We
used Rhat statistics and trace plots to verify convergence. We
considered models with different subsets of fixed and random effects,
and used approximate leave-one-out cross validation to identify the best
model \citep{vehtari2016}.

Finally, while it would theoretically be possible to estimate survival
covariate effects in a joint CMR model that integrates capture histories
across all sites, this was impractical due to computational requirements
of the CMR models (namely, run time and memory). In modeling survival
estimates from site-specific CMR analyses, we are in effect conducting
an among-site meta-analysis to better understand how site, cohort, and
individual level characteristics explain variation in survival after
translocation.

@@

We computed yearly survival probabilities for translocated adults
\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
the posterior distribution of individual state trajectories derived from
CMR models. Although we observed yearly variability in adult survival
within a population, the magnitude of this variability was small
compared to among-population variability
(Fig. 3). Thus, we did not include
yearly within-population variability in adult survival in this analysis.
However, within a population there was substantial yearly variability in
the successful recruitment of adults, greater than what we would expect
from Poisson variability around a mean value. Therefore, we allowed for
yearly variability in the probability of year 1 juvenile survival and
recruitment (additional details provided in \textbf{Estimating model
parameters} below). We also could have included environmental
stochasticity in year 2 juvenile survival and recruitment
\(\sigma_{J_2}\), but our elasticity analysis showed that this parameter
had little effect on host growth rate relative to \(\sigma_{J_1}\)
(Figure~\ref{fig-viability-supp}).

@@

To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
predicted true states for naturally-recruited individuals (1=not
recruited, 2=alive, 3=dead, as described in \emph{CMR model structure}),
then calculated the posterior probability of individuals surviving
between consecutive primary periods, conditional on being alive in the
first primary period (e.g., given a value of 2 (alive) in the first
primary period, how often was the value still 2 (alive) in the next
primary period compared to 3 (dead)?) . This yielded posterior
distributions for survival probabilities between primary periods.
However, because the time interval between primary periods differed, the
survival probabilities between different consecutive primary periods
were not directly comparable. To address this, we converted the survival
probabilities between each consecutive primary period to per day death
rates, propagating the uncertainty from the posterior distributions. We
then took a weighted average of these death rates, weighted by the time
interval between primary periods, to get the average per day death rate
over the entire CMR survey. We converted this per day death rate \(d\)
back to a yearly survival probability using
\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
\(\sigma_{A_T}\) such that our estimates of average yearly survival
probability were comparable between \(\sigma_{A_R}\) and
\(\sigma_{A_T}\).

@@

Finally, we assessed whether our stochastic model could reproduce
observed trajectories of population recovery. We focused on population
70550 because this was our longest CMR time series for a translocated
population and because this population shows evidence of substantial
post-translocation increases in adult abundance associated with
population establishment and recovery. We simulated our model for 16
years, repeating the simulation 50,000 times. For each run and each
year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
Using Approximate Bayesian Computing and rejection sampling
\citep{kosmala2016}, we identified the top 2\% of trajectories (i.e.,
1000 trajectories) that minimized the sum of squared errors between the
observed and predicted data. The yearly \(\sigma_{J_1}\) values
associated with these ``best'' trajectories represented an approximate
posterior distribution \citep{beaumont2010}. Using these best fit
trajectories, we assessed whether our model could qualitatively describe
the patterns of recovery in the observed data for population 70550.

@@

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{study-design}{%
\subparagraph{Study design}\label{study-design}}

We compared frog genomes sampled in naive versus recovering populations.
Comparing populations with different infection histories allowed larger
sample sizes and replication across the landscape. The alternative
approach of comparing samples from the same populations before and after
Bd exposure isn't feasible in this system because Bd arrived in most MYL
frog populations decades ago and population persistence/recovery is rare
and unpredictable. As a result, samples from recovering populations
collected from before and after Bd exposure are not available and are
unlikely to be available in the future.

\hypertarget{go-analysis}{%
\subparagraph{GO analysis}\label{go-analysis}}

In the Results section of the main paper, we describe the stringent set
of outlier SNPs (identified using a Bonferroni-corrected p-value of
0.01). The liberal set, identified using a Bonferroni-corrected p-value
of 0.05, included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct
genes across 16 contigs. Using this liberal set of outlier variants, we
performed a statistical overrepresentation test (Fisher's exact test)
using Blast2GO \citep{conesa2005}. First, we retrieved the BLAST hits
and mapped GO terms for each gene in our targeted transcriptome. Next,
we compared the 30 unique outlier genes from the liberal set of 38
outlier variants to the complete set of genes in our target
transcriptome to determine if any GO biological functions, molecular
functions, or cellular processes were overrepresented. We then repeated
this process for the set of 33 genes located in the
\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.

\hypertarget{genetic-diversity}{%
\subparagraph{Genetic diversity}\label{genetic-diversity}}

To characterize general patterns of genetic diversity between naive and
persistent populations, we calculated individual heterozygosity using
VCFtools \citep{danecek2011}, and compared average heterozygosity using
a Wilcoxon rank-sum test. Additionally, we used VCFtools to calculate
per-population \(\pi\) in 100kb sliding windows along the genome, and in
each of the 10 outlier windows identified in the splined window
analysis. We compared per-population average \(\pi\) values using
Wilcoxon rank-sum tests.

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

\hypertarget{go-analysis-1}{%
\subparagraph{GO analysis}\label{go-analysis-1}}

We found no significantly over-represented GO terms in the liberal set
of outlier regions, or in the set of 33 genes located in the overlapping
\emph{F\textsubscript{ST}} and \(\pi_{diff}\) splined windows.
Therefore\ldots..

\hypertarget{genetic-diversity-1}{%
\subparagraph{Genetic diversity}\label{genetic-diversity-1}}

@@

\caption{\label{fig-survival-postdens}Evidence that Bd load is not an
important predictor of post-translocation frog survival. Depicted
distributions are the estimated posterior density curves and shaded 95\%
uncertainty intervals for the intercept and all predictor variables from
the best model in the between-site meta-analysis. In the Bayesian
framework in which the model was developed, variables are considered
important predictors if the associated uncertainty interval does not
overlap zero (indicated by the dashed red line). Predictor variables
shown on the y-axis are defined as follows: length = frog size, snow\_t
= winter severity in the year of translocation (measured on April 1),
snow\_t1 = winter severity in the year following translocation (measured
on April 1), day = day of year on which a translocation was conducted,
bd\_load = Bd load, elevation = site elevation, order\_first =
within-site translocation order, sex\_male = frog sex, donor\_70567 and
donor\_72996 = donor population.}

@@

\caption{\label{fig-selectionresults}Evidence for selection on
individual variants in recovering MYL frog populations. (A) PCA
calculated from binary SNPs showing the genomic relationship of samples.
Numeric labels and colors match those in
Fig. 1A. Populations 1-7 are \emph{R.
sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B) qqplot
showing observed and expected p-values for 148,307 SNPs and INDELS.
Dashed line shows p-value that identifies outliers. (C) Manhattan plot
showing p-value for each SNP. SNPs are sorted by genomic position and
contigs are sorted by size. Red line shows p-value that identifies
outliers. Outlier SNPs above this threshold are highlighted and labeled.
Bold labels indicate the presence of at least one non-synonymous SNP in
that gene.}

@@

\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
outlier region containing MHC Class I Classical and MHC Class III gene
regions. The plot was created with SimpleSynteny \citep{veltri2016}
using \emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank
accession GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices
indicate the location of SNP outliers in the TCF19 and VARS1 genes. Gap
sizes for each contig representation are labeled.}

@@

\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
heterozygosity for the Bd-naive and recovering populations included in
the frog genomics study. Individual data points are shown on the plot
with their corresponding site number (from
Fig. 1). A Wilcoxon signed-rank test showed
no significant difference between the two groups.}

@@

\caption{\label{fig-boxplot-pi}For each study population in the frog
evolution study (from Fig. 1A), boxplots
showing nucleotide diversity (\(\pi\)) within 100kb sliding windows
along the genome. The y-axis has been log\textsubscript{10}-transformed
for display purposes. Sites are color-coded as ``Bd-naive'' or
``recovering'', and vertical dashed lines show the genetic breaks
between frog populations as described in
Figure~\ref{fig-selectionresults} A). Within the genetic clusters,
mean \(\pi\) is significantly different between sites (p \textless0.001
Wilcoxon rank-sum test).}

>>>>

Swab extracts were analyzed using standard Bd DNA extraction and qPCR
methods (74), and extracts were analyzed singly instead of in triplicate
(87). For analysis of swabs collected during 2005--2014, we used
standards developed from known concentrations of zoospores (74), and
after 2014, we used standards based on single ITS1 PCR amplicons (88).
Based on paired comparisons between samples analyzed using both types of
standards, Bd in the study area has an average of 60 ITS1 copies per
zoospore. To express all qPCR results as the number of ITS1 copies,
starting quantities obtained using the zoospore standard (measured as
``zoospore equivalents'') were multiplied by 60. In addition, all qPCR
quantities (regardless of standard) were multiplied by 80 to account for
the fact that DNA extracts from swabs were diluted 80-fold during
extraction and PCR (33).

@@

We estimated survival and recruitment for each site using open
population CMR models based on (41). For each individual \(i=1, ..., M\)
on each survey \(j=1, ..., n_j\): \(o_{i, j} = 1\) if the individual was
not detected, and \(o_{i, j}=2\) if the individual was detected. Capture
histories of \(M\) individuals are modeled, although only \(N_s\)
individuals were captured. This parameter expanded data augmentation
allows us to capture the possibility that undetected individuals may
have recruited into the adult population (89). Here, \(M\) was chosen to
be three times the number of observed individuals (\(3N_s\)) to be
considerably greater than our prior guess of \(N_s\).

@@

We computed the likelihood of each individual capture history using the
forward algorithm, and we estimated the latent states using the
forward-backward algorithm (41, 90).

All of the code to specify and fit the model in Stan is available in the
open source mrmr package (67).

@@

The objective of this analysis is to describe the influence of site,
cohort, and individual level characteristics on post-translocation frog
survival. By modeling survival estimates obtained from site-specific
mrmr CMR analyses, we are in effect conducting an among-site
meta-analysis. Although it would theoretically be possible to estimate
survival covariate effects in a joint CMR model that integrates capture
histories across all sites, this was impractical due to computational
requirements of the CMR models (namely, run time and memory).

We used Bayesian generalized linear mixed models to investigate
predictors of survival among sites. The response \(y_i\) is binary,
representing a point estimate of whether individual \(i\) survived in
the year following translocation. We generated these point estimates by
rounding the posterior median of 1-year post-introduction survival for
each individual (from site-specific mrmr CMR models) and modeled the
data using a Bernoulli distribution:

@@

These models were fit using the \texttt{stan\_glmer()} function in the
rstanarm package, with default priors described below (71). These priors
are vague, but include data-dependent scaling as follows to account for
different input variable scales. However, because we standardized all
predictor variables similarly to have equal variance (by centering and
dividing by twice the sample standard deviation), the resulting priors
are identical. Specifically, we have:

@@

The default prior for group level adjustments \(\nu_1, ..., \nu_G\) in
rstanarm is a zero-mean Gaussian, where the covariance matrix is
constructed from a correlation matrix with an LKJ prior, and a vector of
variance parameters -- the decomposition of variance prior with unit
regularization, concentration, shape, and scale parameters (91).

We drew posterior samples using Dynamic Hamiltonian Monte Carlo in Stan,
with four parallel chains, each run for 10,000 iterations, discarding
the first half of each chain as warm-up draws (71). We used Rhat
statistics and trace plots to verify convergence. We considered models
with different subsets of fixed and random effects, and used approximate
leave-one-out cross validation to identify the best model (72).

@@

We computed yearly survival probabilities for translocated adults
\(\sigma_{A_T}\) and naturally recruited adults \(\sigma_{A_R}\) from
the posterior distribution of individual state trajectories derived from
mrmr CMR models. Although we observed yearly variability in adult
survival within a population, the magnitude of this variability was
small compared to among-population variability
(Fig. 3). Thus, we did not include
yearly within-population variability in adult survival in this analysis.
However, within a population there was substantial yearly variability in
the successful recruitment of adults, greater than what we would expect
from Poisson variability around a mean value. Therefore, we allowed for
yearly variability in the probability of year 1 juvenile survival and
recruitment (additional details provided in \textbf{Estimating model
parameters} below). We also could have included environmental
stochasticity in year 2 juvenile survival and recruitment
\(\sigma_{J_2}\), but our elasticity analysis (\textbf{Supporting
Information-Population viability modeling-Model analysis and
simulation}) showed that this parameter had little effect on host growth
rate relative to \(\sigma_{J_1}\) (Figure~\ref{fig-viability-supp}).

@@

To estimate the \(\sigma_{A_R}\), we used the posterior distribution of
predicted true states for naturally-recruited individuals (1=not
recruited, 2=alive, 3=dead, as described in \textbf{Supporting
Information-Frog population recovery-CMR model structure}), then
calculated the posterior probability of individuals surviving between
consecutive primary periods, conditional on being alive in the first
primary period (e.g., given a value of 2 (alive) in the first primary
period, how often was the value still 2 (alive) in the next primary
period compared to 3 (dead)?) . This yielded posterior distributions for
survival probabilities between primary periods. However, because the
time interval between primary periods differed, the survival
probabilities between different consecutive primary periods were not
directly comparable. To address this, we converted the survival
probabilities between each consecutive primary period to per day death
rates, propagating the uncertainty from the posterior distributions. We
then took a weighted average of these death rates, weighted by the time
interval between primary periods, to get the average per day death rate
over the entire CMR survey. We converted this per day death rate \(d\)
back to a yearly survival probability using
\(\exp(-d \times 365 \text{ days})\). We used the same procedure for
\(\sigma_{A_T}\) such that our estimates of average yearly survival
probability were comparable between \(\sigma_{A_R}\) and
\(\sigma_{A_T}\).

@@

Finally, we assessed whether our stochastic model could reproduce
observed trajectories of population recovery. We focused on population
70550 because this was our longest CMR time series for a translocated
population and because this population shows evidence of substantial
post-translocation increases in adult abundance associated with
population establishment and recovery. We simulated our model for 16
years, repeating the simulation 50,000 times. For each run and each
year, we drew \(\sigma_{J_1}\) from a uniform distribution between 0 and
1 (or equivalently a beta distribution with mean 0.5 and \(\phi = 2\)).
Using Approximate Bayesian Computing and rejection sampling (92), we
identified the top 2\% of trajectories (i.e., 1000 trajectories) that
minimized the sum of squared errors between the observed and predicted
data. The yearly \(\sigma_{J_1}\) values associated with these ``best''
trajectories represented an approximate posterior distribution (93).
Using these best fit trajectories, we assessed whether our model could
qualitatively describe the patterns of recovery in the observed data for
population 70550.

@@

\hypertarget{study-design}{%
\subsection{Study design}\label{study-design}}

To gain insights into the role of evolution in the development of
resistance by MYL frogs, we compared frog genomes sampled in naive
versus recovering populations. Comparing populations with different
infection histories allowed larger sample sizes and replication across
the landscape. The alternative approach of comparing samples from the
same populations before and after Bd exposure isn't feasible in this
system because Bd arrived in most MYL frog populations decades ago and
population persistence/recovery is rare and unpredictable. As a result,
samples from recovering populations collected from before and after Bd
exposure are not available and are unlikely to be available in the
future.

Our study design, in which we compared frog genomes in naive and
recovering populations, required sampling populations across the range
of MYL frogs in the southern Sierra Nevada. This is due to the fact that
very few MYL frog populations remain in the naive state, and those that
do are scattered across a wide latitudinal range, from Yosemite National
Park in the north to southern Kings Canyon National Park in south. To
minimize potential confounding effects caused by known variation in frog
genotypes across latitude (49), we selected sampling sites such that
both population types were represented across similar latitudinal ranges
(Fig. 1).

\hypertarget{go-analysis}{%
\subsection{GO analysis}\label{go-analysis}}

In \textbf{Results-Frog evolution in response to Bd}, we describe the
stringent set of outlier variants (identified using a
Bonferroni-corrected p-value of 0.01). A liberal set of outlier
variants, identified using a Bonferroni-corrected p-value of 0.05,
included 38 outliers (35 SNPs and 3 INDELS) from 30 distinct genes
across 16 contigs. We used this liberal set to determine if any GO
biological functions, molecular functions, or cellular processes were
overrepresented. To do this, we retrieved the BLAST hits and mapped GO
terms for each gene in our targeted transcriptome. We then conducted a
statistical overrepresentation test (Fisher's exact test) using Blast2GO
(94) to compare the 30 unique outlier genes to the complete set of genes
in our target transcriptome. We repeated this process for the set of 33
genes located in the \emph{F\textsubscript{ST}} and \(\pi_{diff}\)
splined windows.

\hypertarget{genetic-diversity}{%
\subsection{Genetic diversity}\label{genetic-diversity}}

To characterize general patterns of genetic diversity between naive and
persistent populations, we calculated individual heterozygosity using
VCFtools (85), and compared average heterozygosity using a Wilcoxon
rank-sum test. Additionally, we used VCFtools to calculate
per-population \(\pi\) in 100kb sliding windows along the genome, and in
each of the 10 outlier windows identified in the splined window
analysis. We compared per-population average \(\pi\) values using
Wilcoxon rank-sum tests.

@@

\caption{\label{fig-survival-postdens}Results from the among-site
meta-analysis, showing that Bd load is not an important predictor of
post-translocation frog survival. Depicted distributions are the
estimated posterior density curves and shaded 95\% uncertainty intervals
for the intercept and all predictor variables from the best model. In
the Bayesian framework in which the model was developed, variables are
considered important predictors if the associated uncertainty interval
does not overlap zero (indicated by the dashed red line). Predictor
variables shown on the y-axis are defined as follows: length = frog
size, snow\_t = winter severity in the year of translocation (measured
on April 1), snow\_t1 = winter severity in the year following
translocation (measured on April 1), day = day of year on which a
translocation was conducted, bd\_load = Bd load, elevation = site
elevation, order\_first = within-site translocation order, sex\_male =
frog sex, donor\_70567 and donor\_72996 = donor population.}

@@

\caption{\label{fig-selectionresults}Results from the analysis of
individual variants, showing putative signatures of selection in
recovering MYL frog populations. (A) PCA calculated from binary SNPs
showing the genomic relationship of samples. Numeric labels and colors
match those in Fig. 1A. Populations 1-7 are
\emph{R. sierrae} and populations 8 and 9 are \emph{R. muscosa}. (B)
qqplot showing observed and expected p-values for 148,307 SNPs and
INDELS. Dashed line shows p-value that identifies outliers. (C)
Manhattan plot showing p-value for each SNP. SNPs are sorted by genomic
position and contigs are sorted by size. Red line shows p-value that
identifies outliers. Outlier SNPs above this threshold are highlighted
and labeled. Bold labels indicate the presence of at least one
non-synonymous SNP in that gene.}

@@

\caption{\label{fig-synteny-plot}Synteny plot showing conserved gene
order in \emph{Xenopus tropicalis} and \emph{Rana muscosa} for the
outlier region containing MHC Class I Classical and MHC Class III gene
regions. The plot was created with SimpleSynteny (95) using
\emph{Xenopus tropicalis} Chromosome 8 (NC\_030684.2, genbank accession
GCA\_000004195.4) and \emph{Rana muscosa} Contig19. Asterices indicate
the location of SNP outliers in the TCF19 and VARS1 genes. Gap sizes for
each contig representation are labeled.}

@@

\caption{\label{fig-violinplot-heterozy}Violin plots showing individual
heterozygosity for the Bd-naive and recovering populations included in
the frog evolution study. Individual data points are represented by
their corresponding site number (from
Fig. 1). A Wilcoxon signed-rank test showed
no significant difference between the two groups.}

@@

\caption{\label{fig-boxplot-pi}For each study population in the frog
evolution study (from Fig. 1), boxplots
showing nucleotide diversity (\(\pi\)) within 100kb sliding windows
along the genome. The y-axis has been log\textsubscript{10}-transformed
for display purposes. Sites are color-coded as ``Bd-naive'' or
``recovering'', and vertical dashed lines show the genetic breaks
between frog populations as described in
Figure~\ref{fig-selectionresults} A). Within the genetic clusters,
mean \(\pi\) is significantly different between sites (p \textless0.001
Wilcoxon rank-sum test).}
